---
title: "esa_analysis_sans_js"
author: "Bryan Milstead"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(stringsAsFactors = FALSE) 
library(knitr)
library(vegan)
#library(vegan3d)
library(ecodist)
library(readxl)
library(ggplot2)
library(randomForest)
#library(labdsv) #this messes with the tidyverse; call explicitly in the indicator species analysis
library(tidyverse)
library(here)
library(wesanderson)
library(hrbrthemes)  #+ theme_ipsum()
library(viridis)
```

```{r functions, include=FALSE}
#functions to see names in alphabetical order
  names1<-function(df) names(df)[order(names(df))]

#function to select enviro fit results for biplot arrow
    #ef=environmental factor
    #target=selected factors
    #scale=scale factor for arrow
getArrow<-function(ef=ef07js,target=c('microcystin','saxitoxin','cylindrospermopsin'),scale=6){
  scrs <- as.data.frame(scores(ef, display = "vectors")) 
  scrs <- cbind(scrs, ef = rownames(scrs))
  arrow<-filter(scrs,ef%in%target)%>%
      mutate(NMDS1=NMDS1*scale,NMDS2=NMDS2*scale)
  return(arrow)
}

#function to plot Low Medium and High categories with envfit arrows
    #requires the site NMDS scores df "sites"
    #requires a ggplot object with the lake grps='gg'
  #cat=environmental data with low, med, and high categories. e.g., cat=env$mic_tox_cat
  #ef=envirofit df for biplot arrow. e.g., ef=ef07js
  #target=ef arrows to plot. e,g., target=c('mic_det','microcystin')
  #scale= scaling factor for biplot arrows
  #title=title for the plot
  #xnudge=x adjustment for arrow label; note this is a vector with an adjustment level for each arrrow. e.g., xnudge=c(0,0)
  #ynudge=y adjustment for arrow label; note this is a vector with an adjustment level for each arrrow. e.g., ynudge=c(0,0)
  #angle to adjust arrow annotations
  

plotLMH<-function(cat=env$mic_tox_cat,target=c('mic_det','microcystin'),ef=ef07js,scale=7,title='microcystin',xnudge=c(0,0),ynudge=c(0,0),angle=0){
  #combine cat variable, NMDS scores, and size variable for the point
    keep<-cbind(sites,cat=cat)%>%filter(cat!='none')%>%mutate(Size=ifelse(cat=='low',2,ifelse(cat=='med',3,4)))
  #define the biplot arrows
    scrs <- as.data.frame(scores(ef07js, display = "vectors")) 
    scrs <- cbind(scrs, ef = rownames(ef.scrs))
    arrow<-filter(scrs,ef%in%target)%>%
      mutate(NMDS1=NMDS1*scale,NMDS2=NMDS2*scale)
  #plot the cat in NMDS space
    qq<-gg+geom_point(data=keep,aes(x=NMDS1,y=NMDS2,col=factor(cat),size=factor(Size)))+ 
      ggtitle(title)+
      theme(plot.title = element_text(lineheight=4.8, face="bold",hjust = 0.5))+
      #scale_color_manual(values=wes_palette(n=3, name="GrandBudapest"))
      scale_colour_manual(name = "Category\nLevel",
                          labels = c("low","medium","high"),
                          values = viridis_pal()(3)) +   #c("#999999", "#E69F00", "#56B4E9")
      scale_size_manual(name = "Category\nLevel",  
                        labels = c("low","medium","high"),
                        values = c(2,4,6))+
      geom_segment(data =arrow,
                   aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
                   arrow = arrow(length = unit(0.25, "cm")), colour = "blue",lwd=2) +
      geom_text(data = arrow, aes(x = NMDS1, y = NMDS2, label = ef),
                size = 5,nudge_x=xnudge, nudge_y = ynudge, angle=angle) 
    #browser()
    return(qq)
}

#plotLMH()
#plotLMH(cat=env$mic_tox_cat,target=c('mic_det','microcystin'),ef=ef07js,scale=7,title='microcystin',xnudge=c(0,0),ynudge=c(-.1,.1))

#plotLMH(cat=env$cyl_tox_cat,target=c('cyl_det','cylindrospermopsin'),ef=ef07js,scale=7,title='cylindrospermopsin',xnudge=c(0,0),ynudge=c(-.1,.1))

#qq<-plotLMH(cat=env$cyl_prod_abund_cat,target=c('cyl_det','cylindrospermopsin'),ef=ef07js,scale=7,title='cylindrospermopsin',xnudge=c(0,0),ynudge=c(-.2,.1))
#qq


#cat=env$cyl_tox_cat;target=c('cyl_det','cylindrospermopsin');ef=ef07js;scale=7;title='cylindrospermopsin';xnudge=c(0,0);ynudge=c(-.1,.1)
#cat=env$mic_tox_cat;target=c('mic_det','microcystin');ef=ef07js;scale=7;title='microcystin';xnudge=c(0,0);ynudge=c(-.1,.1)

#cat=env$cyano_abund_cat,target=c('cyano_abund','cyl_det','mic_det','sax_det'),ef=ef07js,scale=7,title='cyanobacteria abundance',xnudge=c(0,0,0,0),ynudge=c(.1,.1,0,0)


```


### to do
* 

## Introduction

* modified from Betty_NMSCode_20160824.R

* read taxonomy data
    - mutate: for is.na(genus_new) genus_new=genus (this will update any name changes.)
* read phyto counts (2007 and 2012)
* join taxonomy and counts
* filter
    - !is.na(phylum)-this eliminates data without a valid genus name
    - VISIT_NO==1
    - LAKE_SAMP=='Target_Sampled'
* sum abundance by site_id and genus
* create datasets phyto07 by spreading genus and abund; replace NA with zeros
* create df cyano07 with the cyanobacteria data onl7


```{r data1, include=FALSE, echo=FALSE, cache=FALSE} 
load(file=here('data/phyto07.rda'))
```

* run the NMDS on the phyto abundances with data by taxonomist "JS" removed

```{r NMS07js, include=FALSE, echo=FALSE, cache=FALSE,eval=FALSE} 
nms07js<-metaMDS(phyto07js[,-1],k=4,trymax=10000,parallel=32)
save(nms07js,file=here("data/nms07js.rda"))
```

```{r loadNMS07js, include=FALSE,eval=TRUE}
load(here("data/nms07js.rda"))
```


* load the environmental data (see "esa_analysis.rmd")
* start with the env data from the ecosphere paper
* add Keith Loftins Toxin data
    - create detect fields for each toxin; 1=detect 2=non-detect;cyl_det, mic_det, & sax_det
    - create 'tox_detects' field to indicate the number of toxins detected (0-3)
    - create toxin Low Med High field for each toxin
        - cyl_tox_cat: non=[0,.05); low= [.05,1); med=[1,2); high=[2,inf); NOTE: low  .05<= cyl < 1
        - mic_tox_cat: non=[0,.1); low= [.1,10); med=[10,20); high=[20,inf); Note: this is the who standard
        - sax_tox_cat: non=[0,.02); low= [.02,.1); med=[.1,.2); high=[.2,inf)
* add WHO standards for cyano cells counts
    - cyano_abund_cat: none=[0,1), low= [0,20k); med=[20k,100k); high=[100k,inf)
* add cat for cell counts of possible toxin producers
    - cyl_prod_abund_cat: none=[0,1), low= [0,20k); med=[20k,100k); high=[100k,inf)  
    - mic_prod_abund_cat: none=[0,1), low= [0,20k); med=[20k,100k); high=[100k,inf)
    - sax_prod_abund_cat: none=[0,1), low= [0,20k); med=[20k,100k); high=[100k,inf)
* add NLA chla data (ug/l)
* add WHO standards for Chla (ug/l)
    - chla_cat: low= [0,10) med=[10,50) high=[50,inf)
* add NLA temp_mean
* add "taxonomist"-to see if there is a pattern to who did the classifications


* load the env data and show the data definitions
* filter env sites for env$site_id%in%phyto07js$site_id
* data definitions for the environmental data

```{r env_df, include=TRUE, echo=FALSE}
load(here('data/env.rda'))
env<-filter(env,site_id%in%phyto07js$site_id)
dim(env) #764 95

kable(env_df)

```

* use envfit to compare NMDS scores and environmental gradient data

```{r ev07js, include=FALSE,eval=FALSE}
ef07js<-envfit(nms07js, env, permu = 999,na.rm=TRUE)
save(ef07js,file=here("data/ef07js.rda"))
```

```{r loadEFdata, include=FALSE,eval=TRUE}
load(here("data/ef07js.rda"))
```

* Cluster Analysis

```{r cluster, include=FALSE,eval=FALSE}
dist07<-vegdist(nms07js$points[,1:2],method="euclidean")

clus07com<-hclust(dist07,"complete")
clus07sin<-hclust(dist07,"single")
clus07ave<-hclust(dist07,"average")

cor(dist07, cophenetic(clus07com)) #0.5439064
cor(dist07, cophenetic(clus07sin)) #0.5090613
cor(dist07, cophenetic(clus07ave)) #0.6370945

plot(clus07ave)
rect.hclust(clus07ave,3)

plot(clus07ave)
rect.hclust(clus07ave,4)

grp2<-cutree(clus07ave,2)
grp3<-cutree(clus07ave,3)
grp4<-cutree(clus07ave,4)


table(grp2)
table(grp3)
table(grp4)

save(grp2,grp3,grp4,file=here("data/grps.rda"))

###Plot of NMS color coded for clusters
fig<-ordiplot(nms07js,type="none",main="nms07")
points(fig,"sites",pch=16,cex=0.7,col=grp4)

plot(nms07js,disp='sites',type='n')
ordihull(nms07js,grp4,col=1:4)
ordiellipse(nms07js,grp4,col=1:4,kind='ehull')
ordiellipse(nms07js,grp4,col=1:4,draw="polygon")
ordispider(nms07js,grp4,col=1:4,label=TRUE)
points(nms07js, disp="sites", pch=21, col=1:4, bg=1:4, cex=1.3)
```

```{r cluster, include=FALSE,eval=TRUE}
load(here("data/grps.rda"))
```

* create a table with the 2007 toxic genera
    - NOTE: Oscillatoria includes Planktothrix & Phormidium 
* create table of tox cats

* setup ggplot of convex hull of sites by cluster group 

```{r ggHull, include=FALSE}
#sites from NMS
sites <- as.data.frame(scores(nms07js, display = "sites"))
#add toxin detect data & WSA9
sites<-cbind(sites,select(env,WSA_ECO9,cyl_det,mic_det,sax_det,tox_detects,lake_origin,chla_cat))

#select group
group<-grp4

#add group to sites
sites$group<-group

#define convex hulls by group
hulls <- sites %>% 
  group_by(group) %>% 
  do(.[chull(.[c('NMDS1','NMDS2')]), ])

#create gglot with sites and convex hull
gg<-ggplot(sites) +
  geom_polygon(data = hulls, alpha = 0.3,aes(x=NMDS1,y=NMDS2,fill=factor(group))) +
  coord_fixed() + ## need aspect ratio of 1!
  scale_fill_manual(values=viridis_pal()(max(group)),        
                    name="Assemblage",
                    breaks=unique(group),
                    labels=c('one','two','three','four')[1:max(group)])


```


* plot convex hull of sites by cluster group and points of potentially toxic producing species

```{r pot_tox, include=TRUE, echo=FALSE}
#add toxin producing spp
  #get the toxin producing genera
tox<-read.csv(here("data/toxLoftin.csv"))%>%dplyr::select(-comment,-anatoxin,-nodularin) 

#species
spp <- as.data.frame(scores(nms07js, display = "species"))
spp$genus<-row.names(spp)
spp<-left_join(spp,tox)

#df "tox_prod" has the potential toxin producers
cyl<-filter(spp,cylindrospermopsin==1)%>%mutate(toxin='cylindrospermopsin',size=16)%>%select(NMDS1,NMDS2,toxin,size)
mic<-filter(spp,microcystin==1)%>%mutate(toxin='microcystin',size=8)%>%select(NMDS1,NMDS2,toxin,size)
sax<-filter(spp,saxitoxin==1)%>%mutate(toxin='saxitoxin',size=4)%>%select(NMDS1,NMDS2,toxin,size)
tox_prod<-rbind(cyl,mic,sax)

gg1<-gg +
  geom_point(data=tox_prod,aes(x=NMDS1,y=NMDS2,col=toxin,size=size)) +
  scale_color_manual(values=viridis_pal()(3),                          #c("red", "blue","orange"), 
                  name="Potential\nToxin\nProducer",
                  breaks=c('cylindrospermopsin','microcystin','saxitoxin'),
                  labels=c("cyl", "mic","sax"))+
  scale_size(guide = 'none')

#add names to points
tox_spp<-filter(spp,cylindrospermopsin==1|microcystin==1|saxitoxin==1)
  tox_spp$
    hj<-c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1)
  tox_spp$nx<-c(-.05,-.05,-.05,-.05,-.05,-.05,-.05,-.05,-.05,-.05,-.05,-.05,-.05,-.05,-.05,-.05,-.05)
  tox_spp$ny<-c(0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0)

qq<-gg1+geom_text(data=tox_spp,aes(x=NMDS1,y=NMDS2,label=genus),
                  hjust=c(0,0,0,1,1,
                          1,1,1,0,1,
                          0,1,1,1,1,1,0),
                  nudge_x=c(.03,.02,.05,-.05,-.05,
                            -.05,-.01,-.05,.05,-.03,
                            .05,-.05,-.05,-.05,-.05,-.05,.05),
                  nudge_y=c(-.07,.04,0,0,0.04,
                            -.05,.07,-.03,0,.05,
                            .03,-.03,.05,.07,0,0,-0.05))+
      ggtitle('Potential Toxin Producing Cyanobacteria')+
      theme(plot.title = element_text(lineheight=4.8, face="bold",hjust = 0.5))
qq

select(tox_spp,genus)

###output plot
jpeg(here("output/js_nms_grp_tox_prod_names.jpeg")) 
qq
dev.off()

#arrows and tox producers
#get envfit scores
ef.scrs <- as.data.frame(scores(ef07js, display = "vectors"))
ef.scrs <- cbind(ef.scrs, ef = rownames(ef.scrs))
arrows<-getArrow(ef=ef07js,target=c('microcystin','saxitoxin','cylindrospermopsin'),scale=6)

qq<-gg1 +
  geom_segment(data =arrows,
               aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")), colour = "blue",lwd=2) +
  geom_text(data = arrows, aes(x = NMDS1, y = NMDS2, label = ef),
            size = 5,nudge_x=c(-.2,-.28,0.08), nudge_y = c(.45,-.0,-.23),angle=c(-65,0,-69))+
      ggtitle('Potential Toxin Producing Cyanobacteria')+
      theme(plot.title = element_text(lineheight=4.8, face="bold",hjust = 0.5))
qq

###output plot
jpeg(here("output/js_nms_grp_tox_prod_det_arrow.jpeg")) 
qq
dev.off()
  
```

* plot Low/Med/High cat variables with lake groups and biplot arrows
    - cyl_tox_cat
    - cyl_prod_abund_cat
    - mic_tox_cat
    - mic_prod_abund_cat
    - sax_tox_cat
    - sax_prod_abund_cat
    - cyano_abund_cat
    - chla_cat
    
```{r lmhPlots, include=TRUE, echo=FALSE}

#cyl toxin
  qq<-plotLMH(cat=env$cyl_tox_cat,target=c('cyl_det','cylindrospermopsin','cyl_prod_abund'),ef=ef07js,scale=3,title='Cylindrospermopsin Toxin Category',xnudge=c(-.30,.2,.4),ynudge=c(.42,-.02,-.03),angle=c(-49,0,0))
  qq
  ###output plot
  jpeg(here("output/js_nms_grp_cyl_tox_cat.jpeg")) 
  qq
  dev.off()

#cyl prod
  qq<-plotLMH(cat=env$cyl_prod_abund_cat,target=c('cyl_det','cylindrospermopsin','cyl_prod_abund'),ef=ef07js,scale=3,title='Potential Cylindrospermopsin Producer Abundance',xnudge=c(-.60,.5,.4),ynudge=c(.6,-.5,-.03),angle=c(-49,-45,0))
  qq
  ###output plot
  jpeg(here("output/js_nms_grp_cyl_prod_abund_cat.jpeg")) 
  qq
  dev.off()

#mic toxin
  qq<-plotLMH(cat=env$mic_tox_cat,target=c('mic_det','microcystin','mic_prod_abund'),ef=ef07js,scale=4,title='Microcystin Toxin Category',xnudge=c(-.5,.1,-.2),ynudge=c(-.5,-.05,-.5),angle=c(45,0,65))
  qq
  ###output plot
  jpeg(here("output/js_nms_grp_mic_tox_cat.jpeg")) 
  qq
  dev.off()

#mic prod
  qq<-plotLMH(cat=env$mic_prod_abund_cat,target=c('mic_det','microcystin','mic_prod_abund'),ef=ef07js,scale=4,title='Potential Microcystin Producer Abundance',xnudge=c(-.55,.1,0),ynudge=c(-.6,-.05,-.5),angle=c(45,0,90))
  qq
  ###output plot
  jpeg(here("output/js_nms_grp_mic_prod_abund_cat.jpeg")) 
  qq
  dev.off()  
  
#sax toxin
qq<-plotLMH(cat=env$sax_tox_cat,target=c('sax_det','saxitoxin','sax_prod_abund'),ef=ef07js,scale=3,title='Saxitoxin Toxin Category',xnudge=c(.15,.03,0),ynudge=c(-.2,-.25,-.05),angle=c(-60,-80,0))
qq
###output plot
jpeg(here("output/js_nms_grp_sax_tox_cat.jpeg")) 
qq
dev.off()

#sax prod
  qq<-plotLMH(cat=env$sax_prod_abund_cat,target=c('sax_det','saxitoxin','sax_prod_abund'),ef=ef07js,scale=4,title='Potential Saxitoxin Producer Abundance',xnudge=c(.25,.1,0),ynudge=c(-.5,-.4,-.05),angle=c(-60,-80,0))
  qq
  ###output plot
  jpeg(here("output/js_nms_grp_sax_prod_abund_cat.jpeg")) 
  qq
  dev.off() 
  
#cyano_abund
  qq<-plotLMH(cat=env$cyano_abund_cat,target=c('cyano_abund','cyl_det','mic_det','sax_det'),ef=ef07js,scale=4,title='Cyanobacteria Abundance',xnudge=c(.17,-.21,-.44,.4),ynudge=c(.15,.1,-.1,-.5),angle=c(0,0,0,-33))
  qq
  ###output plot
  jpeg(here("output/js_nms_grp_cyano_abund_cat.jpeg")) 
  qq
  dev.off()

#chla_cat
  qq<-plotLMH(cat=env$chla_cat,target=c('chla','cyl_det','mic_det','sax_det'),ef=ef07js,scale=4,title='Chloropyll a Level',xnudge=c(.15,-.1,-.35,-.15),ynudge=c(.15,.1,-.1,.1))
  qq
  ###output plot
  jpeg(here("output/js_nms_grp_chla_cat.jpeg")) 
  qq
  dev.off()

```

* color lake points by "Community Type"  ##################################

```{r grouping, include=TRUE, echo=FALSE}
#add site points
qq<-gg+geom_point(aes(x=NMDS1,y=NMDS2,col=factor(grp4)),size=2)+
      ggtitle('Lake Clusters')+
      scale_color_manual(values=viridis_pal()(4),                          #c("red", "blue","orange"), 
                  name="Assemblage",
                  breaks=c(1:4),
                  labels=c("one","two","three","four"))+
      theme(plot.title = element_text(lineheight=4.8, face="bold",hjust = 0.5))
qq

###output plot
jpeg(here("output/js_nms_grp_community_type.jpeg")) 
qq
dev.off()
```

* color lake points by "lake_origin"

```{r lake_origin, include=TRUE, echo=FALSE}
#add site points
qq<-gg+geom_point(aes(x=NMDS1,y=NMDS2,col=factor(lake_origin)),size=2)+
      ggtitle('Natural vs. Man-made Lakes')+
      scale_color_manual(values=viridis_pal()(4)[2:3],                          #c("red", "blue","orange"), 
                  name="Origin",
                  breaks=c('MAN-MADE','NATURAL'),
                  labels=c("man-made","natural"))+
      theme(plot.title = element_text(lineheight=4.8, face="bold",hjust = 0.5))
qq

###output plot
jpeg(here("output/js_nms_grp_lake_origen.jpeg")) 
qq
dev.off()
```

* color lake points by "WSA_ECO9"

```{r WSA_ECO9, include=TRUE, echo=FALSE}
#add site points

c('CPL','NAP','NPL','SAP','SPL','TPL','UMW','WMT','XER')
a<-select(env,WSA_ECO9)%>%distinct()%>%arrange(WSA_ECO9)
qq<-gg+geom_point(aes(x=NMDS1,y=NMDS2,col=factor(WSA_ECO9)),size=2)+
      ggtitle('WSA9 Ecoregion')+
      scale_color_manual(values=viridis_pal()(9),                          #c("red", "blue","orange"), 
                  name="Ecoregion",
                  breaks=c('CPL','NAP','NPL','SAP','SPL','TPL','UMW','WMT','XER'),
                  labels=c('CPL','NAP','NPL','SAP','SPL','TPL','UMW','WMT','XER'))+
      theme(plot.title = element_text(lineheight=4.8, face="bold",hjust = 0.5))
qq

###output plot
jpeg(here("output/js_nms_grp_wsa_eco9.jpeg")) 
qq
dev.off()
```

* map community type with biplot arrows for: 

```{r biplot, include=TRUE, echo=FALSE}

#all in one
arrow<-getArrow(ef=ef07js,target=c('mic_det','sax_det','cyl_det','cyano_abund','chla','PTL','NTL'),scale=3)

qq<-gg+geom_segment(data =arrow,
                   aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
                   arrow = arrow(length = unit(0.25, "cm")), colour = "blue",lwd=2) +
      geom_text(data = arrow, aes(x = NMDS1, y = NMDS2, label = ef),
                size = 5,nudge_x=c(0,-.02,0,0,0,-.05,0), nudge_y = c(-.04,-.0,-.06,-.06,-.06,-.0,-.06), angle=c(0,0,0,0,-80,0,0),hjust=c(1,1,0,0,0,1,0)) +
  ggtitle('Biplot Analysis')+theme(plot.title = element_text(lineheight=4.8, face="bold",hjust = 0.5))

qq

###output plot
jpeg(here("output/js_nms_grp_biplot.jpeg")) 
qq
dev.off()

#env biplot
arrow<-getArrow(ef=ef07js,target=c('chla','PTL','NTL','MeanDepthCorrect','temp_mean'),scale=2)
arrow<-mutate(arrow,ef=ifelse(ef=='temp_mean','mean_temperature',ef))
arrow<-mutate(arrow,ef=ifelse(ef=='MeanDepthCorrect','mean_depth',ef))

qq<-gg+geom_segment(data =arrow,
                   aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
                   arrow = arrow(length = unit(0.25, "cm")), colour = "green",lwd=2) +
      geom_text(data = arrow, aes(x = NMDS1, y = NMDS2, label = ef),
                size = 5,nudge_x=c(-.1,-.1,0.35,-.05,-.5), nudge_y = c(.1,-.1,0,.1,0), angle=c(0,0,0,0,0)) +
  ggtitle('Biplot Environmental Variables')+theme(plot.title = element_text(lineheight=4.8, face="bold",hjust = 0.5))

qq

###output plot
jpeg(here("output/js_nms_grp_biplot_env.jpeg")) 
qq
dev.off()

#cyano biplot

arrow<-getArrow(ef=ef07js,target=c('mic_det','sax_det','cyl_det','cyano_abund'),scale=3)

qq<-gg+geom_segment(data =arrow,
                   aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
                   arrow = arrow(length = unit(0.25, "cm")), colour = "blue",lwd=2) +
      geom_text(data = arrow, aes(x = NMDS1, y = NMDS2, label = ef),
                size = 5,nudge_x=c(0,0,0,-.05), nudge_y = c(-.06,-.06,-.06,-.0), angle=c(0,0,-80,0),hjust=c(0,0,0,1)) +
  ggtitle('Biplot Cyanobacteria Variables')+theme(plot.title = element_text(lineheight=4.8, face="bold",hjust = 0.5))

qq

###output plot
jpeg(here("output/js_nms_grp_biplot_cyano.jpeg")) 
qq
dev.off()

```

* draw some box plots based on lake group

```{r boxplot, include=FALSE}

boxplot(log(env$chla)~grp4)
boxplot(log(env$PTL)~grp4)
boxplot(log(env$NTL)~grp4)
boxplot(log(env$mic_prod_abund)~grp4)
boxplot(log(env$cyl_prod_abund)~grp4)
boxplot(log(env$sax_prod_abund)~grp4)
boxplot(log(env$cyano_abund)~grp4)
boxplot(log(env$microcystin)~grp4)

summary(env$microcystin[grp4==1])
summary(env$microcystin[grp4==2])
table(grp4,env$mic_prod_abund_cat)

a<-filter(env,microcystin>0)
summary(a$microcystin[grp4==1])
summary(a$microcystin[grp4==2])

```


* Indicator Species Analysis

```{r ind_spp, include=FALSE,eval=FALSE}
#http://rfunctions.blogspot.com/2013/02/multivariate-analysis-indicator-value.html
#http://biol09.biol.umontreal.ca/Beijing_Linear_model/Indicator_species.pdf

ind_spp<-labdsv::indval(phyto07js[,-1],grp2)

rf <- ind_spp$relfrq;names(rf)<-c('relfreq1','relfreq2')  #[ind_spp$pval<=0.05]
ra <- ind_spp$relabu;names(ra)<-c('relabund1','relabund2') #[ind_spp$pval<=0.05]
gr <- ind_spp$maxcls #[ind_spp$pval<=0.05]
iv <- ind_spp$indcls #[ind_spp$pval<=0.05]
pv <- ind_spp$pval #[ind_spp$pval<=0.05]
fr <- apply(phyto07[,-1]>0, 2, sum) #[ind_spp$pval<=0.05]
indvalsummary <- data.frame(group=gr, indval=iv, pvalue=pv, freq=fr,rf,ra)
indvalsummary <- indvalsummary[order(indvalsummary$group, -indvalsummary$indval),]

indvalsummary <-rownames_to_column(indvalsummary)%>%rename(genus=rowname)
indvalsummary[-1]<-round(indvalsummary[-1],3)

head(indvalsummary)

# Let us see the results
indvalsummary

#what about the toxins
indvalsummary_tox<-inner_join(indvalsummary,tox)

#output the data
write.csv(indvalsummary,here('output/js_indval.csv'),row.names = FALSE)
write.csv(indvalsummary_tox,here('output/js_indval_tox.csv'),row.names = FALSE)

```

* Calc summary stats for the lake groups

```{r ind_spp, include=FALSE}
joy<-cbind(env,grp2)
grp_sum<-group_by(joy,grp2)%>%
  summarise(mean_NTL=mean(NTL, na.rm = TRUE),
            mean_PTL=mean(PTL, na.rm = TRUE),
            mean_chla=mean(chla, na.rm = TRUE),
            mean_temperature=mean(temp_mean, na.rm = TRUE),
            mean_depth=mean(MeanDepthCorrect, na.rm = TRUE),
            mean_microcystin=mean(microcystin, na.rm = TRUE),
            mean_saxitoxin=mean(saxitoxin, na.rm = TRUE),
            mean_cylindrospermopsin=mean(cylindrospermopsin, na.rm = TRUE),
            mean_mic_prod_abund=mean(mic_prod_abund, na.rm = TRUE),
            mean_sax_prod_abund=mean(sax_prod_abund, na.rm = TRUE),
            mean_cyl_prod_abund=mean(cyl_prod_abund, na.rm = TRUE),
            mean_cyano_abund=mean(cyano_abund, na.rm = TRUE))

#function to get the percent detects by group for the toxins
getCnt<-function(q=joy$mic_det){
x<-as.matrix(table(joy$grp2,q))
y<-c(round(100*x[1,2]/(x[1,1]+x[1,2]),1),
round(100*x[2,2]/(x[2,1]+x[2,2]),1))
return(y)}

grp_sum$mic_per_det<-getCnt(joy$mic_det)
grp_sum$sax_per_det<-getCnt(joy$sax_det)
grp_sum$cyl_per_det<-getCnt(joy$cyl_det)

write.csv(grp_sum,here('output/js_grp_sum.csv'),row.names = FALSE)


```
