---
title: "esa_analysis"
author: "Bryan Milstead"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(stringsAsFactors = FALSE) 
library(knitr)
library(vegan)
#library(vegan3d)
library(ecodist)
library(readxl)
library(ggplot2)
library(randomForest)
#library(labdsv) #this messes with the tidyverse; call explicitly in the indicator species analysis
library(tidyverse)
library(here)
library(wesanderson)
```

```{r functions, include=FALSE}
#functions to see names in alphabetical order
  names1<-function(df) names(df)[order(names(df))]

#function to select enviro fit results for biplot arrow
    #ef=environmental factor
    #target=selected factors
    #scale=scale factor for arrow
getArrow<-function(ef=ef07,target=c('microcystin','saxitoxin','cylindrospermopsin'),scale=6){
  scrs <- as.data.frame(scores(ef07, display = "vectors")) 
  scrs <- cbind(scrs, ef = rownames(ef.scrs))
  arrow<-filter(scrs,ef%in%target)%>%
      mutate(NMDS1=NMDS1*scale,NMDS2=NMDS2*scale)
  return(arrow)
}

#function to plot Low Medium and High categories with envfit arrows
    #requires the site NMDS scores df "sites"
    #requires a ggplot object with the lake grps='gg'
  #cat=environmental data with low, med, and high categories. e.g., cat=env$mic_tox_cat
  #ef=envirofit df for biplot arrow. e.g., ef=ef07
  #target=ef arrows to plot. e,g., target=c('mic_det','microcystin')
  #scale= scaling factor for biplot arrows
  #title=title for the plot
  #xnudge=x adjustment for arrow label; note this is a vector with an adjustment level for each arrrow. e.g., xnudge=c(0,0)
  #ynudge=y adjustment for arrow label; note this is a vector with an adjustment level for each arrrow. e.g., ynudge=c(0,0)
  

plotLMH<-function(cat=env$mic_tox_cat,target=c('mic_det','microcystin'),ef=ef07,scale=7,title='microcystin',xnudge=c(0,0),ynudge=c(0,0)){
  #combine cat variable, NMDS scores, and size variable for the point
    keep<-cbind(sites,cat=cat)%>%filter(cat!='none')%>%mutate(Size=ifelse(cat=='low',2,ifelse(cat=='med',3,4)))
  #define the biplot arrows
    scrs <- as.data.frame(scores(ef07, display = "vectors")) 
    scrs <- cbind(scrs, ef = rownames(ef.scrs))
    arrow<-filter(scrs,ef%in%target)%>%
      mutate(NMDS1=NMDS1*scale,NMDS2=NMDS2*scale)
  #plot the cat in NMDS space
    qq<-gg+geom_point(data=keep,aes(x=NMDS1,y=NMDS2,col=factor(cat),size=factor(Size)))+ 
      ggtitle(title)+
      theme(plot.title = element_text(lineheight=4.8, face="bold",hjust = 0.5))+
      #scale_color_manual(values=wes_palette(n=3, name="GrandBudapest"))
      scale_colour_manual(name = "Category\nLevel",
                          labels = c("low","medium","high"),
                          values = c("#999999", "#E69F00", "#56B4E9")) +   
      scale_size_manual(name = "Category\nLevel",  
                        labels = c("low","medium","high"),
                        values = c(2,4,6))+
      geom_segment(data =arrow,
                   aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
                   arrow = arrow(length = unit(0.25, "cm")), colour = "blue",lwd=3) +
      geom_text(data = arrow, aes(x = NMDS1, y = NMDS2, label = ef),
                size = 5,nudge_x=xnudge, nudge_y = ynudge) 
    #browser()
    return(qq)
}

#plotLMH()
#plotLMH(cat=env$mic_tox_cat,target=c('mic_det','microcystin'),ef=ef07,scale=7,title='microcystin',xnudge=c(0,0),ynudge=c(-.1,.1))

#plotLMH(cat=env$cyl_tox_cat,target=c('cyl_det','cylindrospermopsin'),ef=ef07,scale=7,title='cylindrospermopsin',xnudge=c(0,0),ynudge=c(-.1,.1))

qq<-plotLMH(cat=env$cyl_prod_abund_cat,target=c('cyl_det','cylindrospermopsin'),ef=ef07,scale=7,title='cylindrospermopsin',xnudge=c(0,0),ynudge=c(-.2,.1))
qq


cat=env$cyl_tox_cat;target=c('cyl_det','cylindrospermopsin');ef=ef07;scale=7;title='cylindrospermopsin';xnudge=c(0,0);ynudge=c(-.1,.1)
cat=env$mic_tox_cat;target=c('mic_det','microcystin');ef=ef07;scale=7;title='microcystin';xnudge=c(0,0);ynudge=c(-.1,.1)

cat=env$cyano_abund_cat,target=c('cyano_abund','cyl_det','mic_det','sax_det'),ef=ef07,scale=7,title='cyanobacteria abundance',xnudge=c(0,0,0,0),ynudge=c(.1,.1,0,0)


```


### to do
* rf variable importance plots

## Introduction

* modified from Betty_NMSCode_20160824.R

* read taxonomy data
    - mutate: for is.na(genus_new) genus_new=genus (this will update any name changes.)
* read phyto counts (2007 and 2012)
* join taxonomy and counts
* filter
    - !is.na(phylum)-this eliminates data without a valid genus name
    - VISIT_NO==1
    - LAKE_SAMP=='Target_Sampled'
* sum abundance by site_id and genus
* create datasets phyto07 and phyto12 by spreading genus and abund; replace NA with zeros
* create presense absence datasets phyto07pa and phyto12pa replacing values of abund>0 with ones
* create cyano07 and cyano12 by filtering for phylum=='Cyanobacteria'
* create cyano07pa and cyano12pa

```{r data, include=FALSE, echo=FALSE, cache=FALSE} 
#get the taxonomy dat
url <- "https://raw.githubusercontent.com/willbmisled/lakes_database/master/output/nla_phyto_taxonomy.xlsx"
destfile <- "datasets.xlsx"
download.file(url, destfile)
tax<-read_excel(destfile,sheet='nla_phyto_taxonomy')

#tax<-read_excel('C:/bryan/rscripts/lakes_database/output/nla_phyto_taxonomy.xlsx',sheet='nla_phyto_taxonomy')
tax<-mutate(tax,genus_new=ifelse(is.na(genus_new),genus,genus_new))

#get the 2007 count data
counts<-read.csv(url('https://www.epa.gov/sites/production/files/2014-10/nla2007_phytoplankton_softalgaecount_20091023.csv'))
    nrow(counts) #31971
counts<-mutate(counts,GENUS=ifelse(GENUS=='Tetra\xebdron','Tetraëdron',GENUS))  #genus 'Tetraëdron' read incorrectly

#merge tax and counts by genus and filter
t2007<-left_join(counts,tax,by=c("GENUS"="genus"))%>%filter(!is.na(phylum),VISIT_NO==1,LAKE_SAMP=='Target_Sampled')%>%rename(genus=genus_new)
    nrow(t2007) #26382

t2007 <- group_by(t2007,SITE_ID,phylum,genus)%>%
  summarise(abund=sum(ABUND, na.rm = TRUE))
    names(t2007)<-tolower(names(t2007))
        nrow(t2007) #24107
t2007<-ungroup(t2007)

#create phyto07  all phytoplankton
phyto07<-select(t2007,-phylum)%>%spread(genus,abund) #1,148 × 255
phyto07[is.na(phyto07)] <- 0 

#create phyto07pa (presence/absence data)
phyto07pa<-phyto07
phyto07pa[,-1][phyto07pa[,-1]>0] <- 1


#create cyano07 cyanobacteria only 
cyano07<-filter(t2007,phylum=='Cyanobacteria')%>%select(-phylum)%>%spread(genus,abund) ##1,115 × 51
cyano07[is.na(cyano07)] <- 0

#create cyano07pa (presence/absence data)
cyano07pa<-cyano07
cyano07pa[,-1][cyano07pa[,-1]>0] <- 1

###2012
#get the 2012 count data
counts<-read.csv(url('https://www.epa.gov/sites/production/files/2017-02/nla2012_wide_phytoplankton_count_02122014.csv'))
counts<-mutate(counts,genus=ifelse(is.na(GENUS),NA,paste(toupper(substr(GENUS,1,1)),tolower(substr(GENUS,2,111)),sep="")))

#merge tax and counts by genus and filter
t2012<-left_join(counts,tax,by="genus")%>%filter(!is.na(phylum),VISIT_NO==1)%>%select(-genus)%>%rename(genus=genus_new)
    nrow(t2012) #33376
    
t2012 <- group_by(t2012,SITE_ID,phylum,genus)%>%
  summarise(abund=sum(ABUNDANCE, na.rm = TRUE))
    names(t2012)<-tolower(names(t2012))
        nrow(t2012) #22017
t2012<-ungroup(t2012)

#create phyto12  all phytoplankton
phyto12<-select(t2012,-phylum)%>%spread(genus,abund) #1,133 × 269
phyto12[is.na(phyto12)] <- 0 

#create phyto12pa (presence/absence data)
phyto12pa<-phyto12
phyto12pa[,-1][phyto12pa[,-1]>0] <- 1

#create cyano12 cyanobacteria only 
cyano12<-filter(t2012,phylum=='Cyanobacteria')%>%select(-phylum)%>%spread(genus,abund) #1,115 × 55
cyano12[is.na(cyano12)] <- 0

#create cyano12pa (presence/absence data)
cyano12pa<-cyano12
cyano12pa[,-1][cyano12pa[,-1]>0] <- 1

```

* decide how many dimensions to use for the NMS

```{r stressTest, include=FALSE, echo=FALSE, cache=FALSE,eval=FALSE} 
#Running NMS fit
phytoNMS.2<-metaMDS(phyto07[,-1],k=2,trymax=500)
phytoNMS.3<-metaMDS(phyto07[,-1],k=3,trymax=500)
phytoNMS.4<-metaMDS(phyto07[,-1],k=4,trymax=500)
phytoNMS.5<-metaMDS(phyto07[,-1],k=5,trymax=500)
phytoNMS.6<-metaMDS(phyto07[,-1],k=6,trymax=500)

#save(phytoNMS.2,phytoNMS.3,phytoNMS.4,file=here("data/phytoNMS_stress_test1.rda"))
#save(phytoNMS.5,phytoNMS.6,file=here("data/phytoNMS_stress_test2.rda"))
#load(file=here("phytoNMS.rda"))

stress<-data.frame(dimension=2:6,stress=c(phytoNMS.2$stress,phytoNMS.3$stress,phytoNMS.4$stress,phytoNMS.5$stress,phytoNMS.6$stress))

plot(stress$dimension,stress$stress)

stressplot(phytoNMS.2)
```

* run the NMDS on the phyto abundances for 2007 and 2012

```{r NMDS07, include=FALSE, echo=FALSE, cache=FALSE,eval=FALSE} 
nms07<-metaMDS(phyto07[,-1],k=4,trymax=10000)
save(nms07,file=here("data/nms07.rda"))
```

```{r NMDS12, include=FALSE, echo=FALSE, cache=FALSE,eval=FALSE} 
nms12<-metaMDS(phyto12[,-1],k=4,trymax=10000)
save(nms12,file=here('data/nms12.rda'))
```

* run the NMDS on phyto presence/absence for 2007 and 2012

```{r NMDS07pa, include=FALSE, echo=FALSE, cache=FALSE,eval=FALSE} 
nms07pa<-metaMDS(phyto07pa[,-1],k=4,trymax=10000)
save(nms07pa,file=here("data/nms07pa.rda"))
```

```{r NMDS12pa, include=FALSE, echo=FALSE, cache=FALSE,eval=FALSE} 
nms12pa<-metaMDS(phyto12pa[,-1],k=4,trymax=10000)
save(nms12pa,file=here("data/nms12pa.rda"))
```

* run the NMDS on the cyano abundances for 2007 and 2012

```{r NMDS07c, include=FALSE, echo=FALSE, cache=FALSE,eval=FALSE} 
nms07c<-metaMDS(cyano07[,-1],k=4,trymax=10000)
save(nms07c,file=here("data/nms07c.rda"))
```

```{r NMDS12c, include=FALSE, echo=FALSE, cache=FALSE,eval=FALSE} 
nms12c<-metaMDS(cyano12[,-1],k=4,trymax=10000)
save(nms12c,file=here('data/nms12c.rda'))
```

* run the NMDS on cyano presence/absence for 2007 and 2012

```{r NMDS07pac, include=FALSE, echo=FALSE, cache=FALSE,eval=FALSE} 
nms07pac<-metaMDS(cyano07pa[,-1],k=4,trymax=10000)
save(nms07pac,file=here("data/nms07pac.rda"))
```

```{r NMDS12pac, include=FALSE, echo=FALSE, cache=FALSE,eval=FALSE} 
nms12pac<-metaMDS(cyano12pa[,-1],k=4,trymax=10000)
save(nms12pac,file=here("data/nms12pac.rda"))
```
* get the environmental data
* start with the env data from the ecosphere paper
* add Keith Loftins Toxin data
    - create detect fields for each toxin; 1=detect 2=non-detect;cyl_det, mic_det, & sax_det
    - create 'tox_detects' field to indicate the number of toxins detected (0-3)
    - create toxin Low Med High field for each toxin
        - cyl_tox_cat: non=[0,.05); low= [.05,1); med=[1,2); high=[2,inf); NOTE: low  .05<= cyl < 1
        - mic_tox_cat: non=[0,.1); low= [.1,10); med=[10,20); high=[20,inf); Note: this is the who standard
        - sax_tox_cat: non=[0,.02); low= [.02,.1); med=[.1,.2); high=[.2,inf)
* add WHO standards for cyano cells counts
    - cyano_abund_cat: none=[0,1), low= [0,20k); med=[20k,100k); high=[100k,inf)
* add cat for cell counts of possible toxin producers
    - cyl_prod_abund_cat: none=[0,1), low= [0,20k); med=[20k,100k); high=[100k,inf)  
    - mic_prod_abund_cat: none=[0,1), low= [0,20k); med=[20k,100k); high=[100k,inf)
    - sax_prod_abund_cat: none=[0,1), low= [0,20k); med=[20k,100k); high=[100k,inf)
* add NLA chla data (ug/l)
* add WHO standards for Chla (ug/l)
    - chla_cat: low= [0,10) med=[10,50) high=[50,inf)

```{r evdata, include=FALSE}
#load the data and data definiitions directly from github fir the ecosphere paper
env<-read.csv(url("https://raw.githubusercontent.com/USEPA/LakeTrophicModelling/master/inst/extdata/ltmData.csv"))
env_df<-read.csv(url("https://raw.githubusercontent.com/USEPA/LakeTrophicModelling/master/inst/extdata/data_def.csv"))

#change NLA_ID to site_ID
env<-rename(env,site_id=NLA_ID)

#add site_id to data def
env_df<-rbind(data.frame(variable_names='site_id',description="unique NLA site id for lake",type='id'),env_df)

#keep only the variables in the data definitions
env<-select_(env,.dots=env_df$variable_names)

#add the algal toxins data
#get raw data
nla2007<-read.csv(url('https://raw.githubusercontent.com/willbmisled/lakes_database/master/output/nla2007_chem.csv'))
#keep the toxin data only and switch to wide format
tox<-filter(nla2007,parameter%in%c("cylindrospermopsin","microcystin","saxitoxin"),duplicate=='P')%>%select(uid,parameter,result)%>%spread(parameter,result)

#add the NLA id
uid<-read.csv(url('https://raw.githubusercontent.com/willbmisled/lakes_database/master/output/nla_samples.csv'))%>%filter(visit_no==1)%>%select(uid,site_id)

tox<-left_join(tox,uid)%>%select(-uid)

#add tox to env
dim(env) #1148 72
env<-left_join(env,tox)
dim(env) #1148 75

#replace missing tox values with zeros
env<-mutate(env,cylindrospermopsin=ifelse(is.na(cylindrospermopsin),0,cylindrospermopsin))
env<-mutate(env,microcystin=ifelse(is.na(microcystin),0,microcystin))
env<-mutate(env,saxitoxin=ifelse(is.na(saxitoxin),0,saxitoxin))

#add toxin detect variables
env<-mutate(env,cyl_det=ifelse(cylindrospermopsin>0,1,0),mic_det=ifelse(microcystin>0,1,0),sax_det=ifelse(saxitoxin>0,1,0))

#add tox_detects (total number of toxins detected)
env<-mutate(env,tox_detects=cyl_det+mic_det+sax_det)

#create toxin Low Med High field for each toxin
env$cyl_tox_cat<-cut(env$cylindrospermopsin,breaks=c(0,.05,1,2,max(env$cylindrospermopsin,na.rm=TRUE)+1),right=FALSE,labels=c('none','low','med','high'),ordered_result=TRUE)

env$mic_tox_cat<-cut(env$microcystin,breaks=c(0,.1,10,20,max(env$microcystin,na.rm=TRUE)+1),right=FALSE,labels=c('none','low','med','high'),ordered_result=TRUE)

env$sax_tox_cat<-cut(env$saxitoxin,breaks=c(0,.02,.1,.2,max(env$saxitoxin,na.rm=TRUE)+1),right=FALSE,labels=c('none','low','med','high'),ordered_result=TRUE)
                 
#add toxins to data def
tox_df<-data.frame(variable_names=c('cylindrospermopsin','microcystin','saxitoxin','cyl_det','mic_det','sax_det','tox_detects','cyl_tox_cat','mic_tox_cat','sax_tox_cat'),
        description=c("Total cylindrospermopsin conc ELISA (µg/L)","Total microcystin conc ELISA (µg/L)","Total saxitoxin conc ELISA (µg/L)","1=cylindrospermopsin conc > reporting limit of 0.05 µg/L","1=microcystin conc > reporting limit of 0.1 µg/L","1=saxitoxin conc > reporting limit of 0.02 µg/L","number of toxins detected","non=[0,.05); low= [.05,1); med=[1,2); high=[2,inf)","non=[0,.1); low= [.1,10); med=[10,20); high=[20,inf); Note: this is the who standard","non=[0,.02); low= [.02,.1); med=[.1,.2); high=[.2,inf)"),
        type=rep('toxin',10))

env_df<-rbind(env_df,tox_df)

dim(env) #1148   82

# add WHO standards for cyano cells counts
c_count<-data.frame(site_id=cyano07$site_id,cyano_abund=apply(cyano07[,-1],1,sum))

c_count$cyano_abund_cat<-cut(c_count$cyano_abund,breaks=c(0,1,20000,100000,max(c_count$cyano_abund,na.rm=TRUE)+1),right=FALSE,labels=c('none','low','med','high'),ordered_result=TRUE)

env<-left_join(env,c_count)

dim(env) #1148   84

env_df<-rbind(data.frame(variable_names=c("cyano_abund","cyano_abund_cat"),description=c("cyano cell counts","WHO cyano category: none=[0,1), low= [0,20k); med=[20k,100k); high=[100k,inf)"),type=rep('cyano',2)),env_df)

# add cat for cell counts of possible toxin producers
    # cyl_prod_abund_cat: none=[0,1), low= [0,20k); med=[20k,100k); high=[100k,inf)  

#get the toxin producing genera
tox<-read_excel(destfile,sheet='toxLoftin')%>%select(-comment,-anatoxin,-nodularin) 

cyl<-filter(tox,cylindrospermopsin==1)
cyl_prod<-select_(cyano07,.dots=c('site_id',names(cyano07)[names(cyano07)%in%cyl$genus]))
cyl_prod$cyl_prod_abund<-apply(cyl_prod[,-1],1,sum)
cyl_prod$cyl_prod_abund_cat<-cut(cyl_prod$cyl_prod_abund,breaks=c(0,1,20000,100000,max(cyl_prod$cyl_prod_abund,na.rm=TRUE)+1),right=FALSE,labels=c('none','low','med','high'),ordered_result=TRUE)
cyl_prod$cyl_prod_abund_cat<-cyl_prod$cyl_prod_abund_cat
env<-left_join(env,select(cyl_prod,site_id,cyl_prod_abund,cyl_prod_abund_cat))

dim(env) #1148   86

mic<-filter(tox,microcystin==1)
mic_prod<-select_(cyano07,.dots=c('site_id',names(cyano07)[names(cyano07)%in%mic$genus]))
mic_prod$mic_prod_abund<-apply(mic_prod[,-1],1,sum)
mic_prod$mic_prod_abund_cat<-cut(mic_prod$mic_prod_abund,breaks=c(0,1,20000,100000,max(mic_prod$mic_prod_abund,na.rm=TRUE)+1),right=FALSE,labels=c('none','low','med','high'),ordered_result=TRUE)
mic_prod$mic_prod_abund_cat<-mic_prod$mic_prod_abund_cat
env<-left_join(env,select(mic_prod,site_id,mic_prod_abund,mic_prod_abund_cat))

dim(env) #1148   88

sax<-filter(tox,saxitoxin==1)
sax_prod<-select_(cyano07,.dots=c('site_id',names(cyano07)[names(cyano07)%in%sax$genus]))
sax_prod$sax_prod_abund<-apply(sax_prod[,-1],1,sum)
sax_prod$sax_prod_abund_cat<-cut(sax_prod$sax_prod_abund,breaks=c(0,1,20000,100000,max(sax_prod$sax_prod_abund,na.rm=TRUE)+1),right=FALSE,labels=c('none','low','med','high'),ordered_result=TRUE)
sax_prod$sax_prod_abund_cat<-sax_prod$sax_prod_abund_cat
env<-left_join(env,select(sax_prod,site_id,sax_prod_abund,sax_prod_abund_cat))


dim(env) #1148   90

#add tox_prod field to data def
tox_df<-data.frame(variable_names=c('cyl_prod_abund','cyl_prod_abund_cat',
                                    'mic_prod_abund','mic_prod_abund_cat',
                                    'sax_prod_abund','sax_prod_abund_cat'),
        description=c('potential cylindrospermopsin producer cell count','potential cylindrospermopsin producer category: none=[0,1), low= [0,20k); med=[20k,100k); high=[100k,inf))','potential microcystin producer cell count','potential microcystin producer category: none=[0,1), low= [0,20k); med=[20k,100k); high=[100k,inf))','potential saxitoxin producer cell count','potential saxitoxin producer category: none=[0,1), low= [0,20k); med=[20k,100k); high=[100k,inf))'),
        type=rep('tox_prod',6))

env_df<-rbind(env_df,tox_df)

# add WHO standards for Chla (ug/l): chla_cat: low= [0,10) med=[10,50) high=[50,inf)
chla<-filter(nla2007,parameter=='chla',duplicate=='P')%>%select(uid,parameter,result)%>%spread(parameter,result)
chla<-left_join(chla,uid)%>%select(-uid) #add the NLA id

chla$chla_cat<-cut(chla$chla,breaks=c(0,10,50,max(chla$chla,na.rm=TRUE)+1),right=FALSE,labels=c('low','med','high'),ordered_result=TRUE)

env<-left_join(env,chla)

dim(env) #1148   92

env_df<-rbind(env_df,data.frame(variable_names=c('chla','chla_cat'),description=c('NLA measured chla ug/l','WHO categories for chla: low= [0,10) med=[10,50) high=[50,inf)'),type=c('NLA','chla_cat')))

# add lake origin from site info file
site<-read.csv(url('https://www.epa.gov/sites/production/files/2014-01/nla2007_sampledlakeinformation_20091113.csv'))
site<-filter(site,VISIT_NO==1,LAKE_SAMP=='Target_Sampled')%>%select(site_id=SITE_ID,lake_origin=LAKE_ORIGIN)
env_df<-rbind(env_df,data.frame(variable_names='lake_origin',description='lake origin; man-made or natural',type='site'))

env<-left_join(env,site)

#make sure env$site_id is in the same order as the nmds data
env<-left_join(select(phyto07,site_id),env) 
table(phyto07$site_id==env$site_id) #check for site_id match-all TRUE

dim(env) #1148   93

table(names(env)%in%env_df$variable_names) #t=93
table(env_df$variable_names%in%names(env)) #t=93
```

* data definitions for the environmental data

```{r env_df, include=TRUE, echo=FALSE}
kable(env_df)

```

```{r loadNMSdata, include=FALSE,eval=TRUE}
load(here("data/nms07.rda"))
```

* use envfit to compare NMDS scores and environmental gradient data

```{r ev7, include=FALSE,eval=FALSE}
ef07<-envfit(nms07, env, permu = 999,na.rm=TRUE)
save(ef07,file=here("data/ef07.rda"))
```

```{r loadEFdata, include=FALSE,eval=TRUE}
load(here("data/ef07.rda"))
```

* Cluster Analysis

```{r cluster, include=FALSE}
dist07<-vegdist(nms07$points[,1:2],method="euclidean")

clus07com<-hclust(dist07,"complete")
clus07sin<-hclust(dist07,"single")
clus07ave<-hclust(dist07,"average")

cor(dist07, cophenetic(clus07com)) #0.6585126
cor(dist07, cophenetic(clus07sin)) #0.5396457
cor(dist07, cophenetic(clus07ave)) #0.7734899

plot(clus07ave)
rect.hclust(clus07ave,3)

grp2<-cutree(clus07ave,2)
grp3<-cutree(clus07ave,3)
grp4<-cutree(clus07ave,4)


table(grp2)
table(grp3)
table(grp4)

###Plot of NMS color coded for clusters
fig<-ordiplot(nms07,type="none",main="nms07")
points(fig,"sites",pch=16,cex=0.7,col=grp3)

plot(nms07,disp='sites',type='n')
ordihull(nms07,grp2,col=1:3)
ordiellipse(nms07,grp2,col=1:3,kind='ehull')
ordiellipse(nms07,grp2,col=1:3,draw="polygon")
ordispider(nms07,grp2,col=1:3,label=TRUE)
points(nms07, disp="sites", pch=21, col="red", bg="yellow", cex=1.3)

```

* create a table with the 2007 toxic genera
    - NOTE: Oscillatoria includes Planktothrix & Phormidium 

```{r toxTable, include=FALSE, echo=FALSE, cache=FALSE,eval=FALSE} 
a<-group_by(t2007,genus)%>%
    summarise(count = n(),
              median= round(median(abund, na.rm = TRUE)),
              mean= round(mean(abund, na.rm = TRUE)),
              max =round(sd(abund, na.rm = TRUE)))%>%
    inner_join(tox)%>%arrange(genus)%>%
      filter(cylindrospermopsin==1|microcystin==1|saxitoxin==1)

write.csv(a,file=here("data/tox_table.csv"),row.names = FALSE)
```

* setup ggplot of convex hull of sites by cluster group 

```{r ggHull, include=FALSE}
#sites from NMS
sites <- as.data.frame(scores(nms07, display = "sites"))
#add toxin detect data & WSA9
sites<-cbind(sites,select(env,WSA_ECO9,cyl_det,mic_det,sax_det,tox_detects,lake_origin,chla_cat))

#select group
group<-grp2

#add group to sites
sites$group<-group

#define convex hulls by group
hulls <- sites %>% 
  group_by(group) %>% 
  do(.[chull(.[c('NMDS1','NMDS2')]), ])


#create gglot with sites and convex hull
gg<-ggplot(sites) +
  geom_polygon(data = hulls, alpha = 0.3,aes(x=NMDS1,y=NMDS2,fill=factor(group))) +
  coord_fixed() + ## need aspect ratio of 1!
  scale_fill_manual(values=c("#E69F00", "#56B4E9"), 
                    name="Lake\nGroup",
                    breaks=c(1, 2),
                    labels=c("One", "Two"))
```


* plot convex hull of sites by cluster group and points of potentially toxic producing species

```{r pot_tox, include=TRUE, echo=FALSE}
#add toxin producing spp
  #get the toxin producing genera
tox<-read_excel(destfile,sheet='toxLoftin')%>%dplyr::select(-comment,-anatoxin,-nodularin) 

#species
spp <- as.data.frame(scores(nms07, display = "species"))
spp$genus<-row.names(spp)
spp<-left_join(spp,tox)

#df "tox_prod" has the potential toxin producers
cyl<-filter(spp,cylindrospermopsin==1)%>%mutate(toxin='cylindrospermopsin',size=6)%>%select(NMDS1,NMDS2,toxin,size)
mic<-filter(spp,microcystin==1)%>%mutate(toxin='microcystin',size=4)%>%select(NMDS1,NMDS2,toxin,size)
sax<-filter(spp,saxitoxin==1)%>%mutate(toxin='saxitoxin',size=2)%>%select(NMDS1,NMDS2,toxin,size)
tox_prod<-rbind(cyl,mic,sax)

gg1<-gg +
  geom_point(data=tox_prod,aes(x=NMDS1,y=NMDS2,col=toxin,size=size)) +
  scale_color_manual(values=c("red", "blue","orange"), 
                  name="Potential\nToxin\nProducer",
                  breaks=c('cylindrospermopsin','microcystin','saxitoxin'),
                  labels=c("cyl", "mic","sax"))+
  scale_size(guide = 'none')

#add names to points
tox_spp<-filter(spp,cylindrospermopsin==1|microcystin==1|saxitoxin==1)

qq<-gg1+geom_text(data=tox_spp,aes(x=NMDS1,y=NMDS2,label=genus),hjust=0,nudge_x=.05)
qq

###output plot
jpeg(here("output/nms_grp_tox_prod_names.jpeg")) 
qq
dev.off()

#arrows and tox producers
arrows<-getArrow(ef=ef07,target=c('microcystin','saxitoxin','cylindrospermopsin'),scale=6)

qq<-gg1 +
  geom_segment(data =arrows,
               aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")), colour = "red",lwd=3) +
  geom_text(data = arrows, aes(x = NMDS1, y = NMDS2, label = ef),
            size = 5)
qq

###output plot
jpeg(here("output/nms_grp_tox_prod_det_arrow.jpeg")) 
qq
dev.off()
  
```

* plot Low/Med/High cat variables with lake groups and biplot arrows
    - cyl_tox_cat
    - cyl_prod_abund_cat
    - mic_tox_cat
    - mic_prod_abund_cat
    - sax_tox_cat
    - sax_prod_abund_cat
    - cyano_abund_cat
    - chla_cat
    
```{r lmhPlots, include=TRUE, echo=FALSE}

#cyl toxin
qq<-plotLMH(cat=env$cyl_tox_cat,target=c('cyl_det','cylindrospermopsin'),ef=ef07,scale=7,title='Cylindrospermopsin Toxin Category',xnudge=c(0,0),ynudge=c(-.2,.1))
qq
###output plot
jpeg(here("output/nms_grp_cyl_tox_cat.jpeg")) 
qq
dev.off()

qq<-plotLMH(cat=env$cyl_prod_abund_cat,target=c('cyl_det','cylindrospermopsin'),ef=ef07,scale=7,title='Potential Cylindrospermopsin Producer Abundance',xnudge=c(0,0),ynudge=c(-.2,.1))
qq
###output plot
jpeg(here("output/nms_grp_cyl_prod_abund_cat.jpeg")) 
qq
dev.off()

qq<-plotLMH(cat=env$mic_tox_cat,target=c('mic_det','microcystin'),ef=ef07,scale=7,title='microcystin',xnudge=c(0,0),ynudge=c(-.1,.1))
qq
###output plot
jpeg(here("output/nms_grp_mic_tox_cat.jpeg")) 
qq
dev.off()

qq<-plotLMH(cat=env$sax_tox_cat,target=c('sax_det','saxitoxin'),ef=ef07,scale=7,title='saxitoxin',xnudge=c(0,0),ynudge=c(.1,.1))
qq
###output plot
jpeg(here("output/nms_grp_sax_tox_cat.jpeg")) 
qq
dev.off()

qq<-plotLMH(cat=env$cyano_abund_cat,target=c('cyano_abund','cyl_det','mic_det','sax_det'),ef=ef07,scale=7,title='cyanobacteria abundance',xnudge=c(0,0,0,0),ynudge=c(.1,.1,0,0))
qq
###output plot
jpeg(here("output/nms_grp_cyano_abund_cat.jpeg")) 
qq
dev.off()

```


* color lake points by "lake_origin"

```{r lake_origin, include=TRUE, echo=FALSE}
#add site points
qq<-gg+geom_point(aes(x=NMDS1,y=NMDS2,col=factor(lake_origin)),size=1) #note size outside aes so no legend
qq

###output plot
jpeg(here("output/nms_grp_lake_origen.jpeg")) 
qq
dev.off()
```

* color lake points by "WSA_ECO9"

```{r WSA_ECO9, include=TRUE, echo=FALSE}
#add site points
qq<-gg+geom_point(aes(x=NMDS1,y=NMDS2,col=factor(WSA_ECO9)),size=1) #note size outside aes so no legend
qq

###output plot
jpeg(here("output/nms_grp_wsa_eco9.jpeg")) 
qq
dev.off()
```

* color lake points by "chla_cat"

```{r chla_ca, include=TRUE, echo=FALSE}
#add site points
gg+geom_point(aes(x=NMDS1,y=NMDS2,col=factor(chla_cat)),size=1) #note size outside aes so no legend
qq

###output plot
jpeg(here("output/nms_grp_chla_cat.jpeg")) 
qq
dev.off()
```


* color lake points by "mic_tox_cat"  #NOTE: this one works convert to function

```{r mic_tox_cat, include=TRUE, echo=FALSE}
#keep only sites with toxin producers
keep<-cbind(sites,mic_tox_cat=env$mic_tox_cat)%>%filter(mic_tox_cat!='non')%>%mutate(size=ifelse(mic_tox_cat=='low',2,ifelse(mic_tox_cat=='med',3,4)))

#add site points& arrows
arrows<-getArrow(ef=ef07,target=c('mic_det','microcystin'),scale=7)

qq<-gg+geom_point(data=keep,aes(x=NMDS1,y=NMDS2,col=factor(mic_tox_cat),size=factor(size)))+ 
      ggtitle("microcystin")+
      theme(plot.title = element_text(lineheight=4.8, face="bold",hjust = 0.5))+
    #scale_color_manual(values=wes_palette(n=3, name="GrandBudapest"))
    scale_colour_manual(name = "toxin\nlevel",
                      labels = c("low","medium","high"),
                      values = c("#999999", "#E69F00", "#56B4E9")) +   
    scale_size_manual(name = "microcystin\nlevel",
                     labels = c("low","medium","high"),
                     values = c(2,4,6))+
      geom_segment(data =arrows,
               aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")), colour = "blue",lwd=3) +
      geom_text(data = arrows, aes(x = NMDS1, y = NMDS2, label = c("mic_Level","mic_Detects")),
            size = 5,nudge_x=c(0,0), nudge_y = c(-.1,.1)) 
qq


###output plot
jpeg(here("output/nms_grp_mic_tox_cat.jpeg")) 
qq
dev.off()
```

* color lake points by "cyl_tox_cat"

```{r cyl_tox_cat, include=TRUE, echo=FALSE}
#keep only sites with toxin producers
keep<-cbind(sites,cyl_tox_cat=env$cyl_tox_cat)%>%filter(cyl_tox_cat!='non')

#add site points& arrows
arrows<-getArrow(ef=ef07,target=c('cyl_det','cylindrospermopsin'),scale=6)

qq<-gg+geom_point(data=keep,aes(x=NMDS1,y=NMDS2,col=factor(cyl_tox_cat)),size=2)+ 
      ggtitle("Cylindrospermopsin")+
      theme(plot.title = element_text(lineheight=4.8, face="bold",hjust = 0.5))+
          scale_colour_manual(name = "Toxin\nLevel",
                      labels = c("Low","Medium","High"),
                      values = c("#999999", "#E69F00", "#56B4E9"))+
      geom_segment(data =arrows,
               aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")), colour = "blue",lwd=3) +
      geom_text(data = arrows, aes(x = NMDS1, y = NMDS2, label = c("Cyl_Level","Cyl_Detects")),
            size = 5,nudge_x=c(0,0), nudge_y = c(-.1,.1)) 
  qq
###output plot
jpeg(here("output/nms_grp_cyl_tox_cat.jpeg")) 
qq
dev.off()
```

* color lake points by "sax_tox_cat"

```{r sax_tox_cat, include=TRUE, echo=FALSE}
#keep only sites with toxin producers
keep<-cbind(sites,sax_tox_cat=env$sax_tox_cat)%>%filter(sax_tox_cat!='non')

#add site points
qq<-gg+geom_point(data=keep,aes(x=NMDS1,y=NMDS2,col=factor(sax_tox_cat)),size=2) #note size outside aes so no legend
qq

###output plot
jpeg(here("output/nms_grp_sax_tox_cat.jpeg")) 
qq
dev.off()
```

* color lake points by "mic_prod_abund_cat"

```{r mic_prod_abund_cat, include=TRUE, echo=FALSE}
#keep only sites with toxin producers
keep<-cbind(sites,mic_prod_abund_cat=env$mic_prod_abund_cat)%>%filter(mic_prod_abund_cat!='non')

#define arrow
scale<-6 
arrow<-filter(ef.scrs,ef=='mic_prod_abund_cat')%>%
      mutate(NMDS1=NMDS1*scale,NMDS2=NMDS2*scale)

#add arrows to ggplot
gg +
  geom_segment(data =arrow,
               aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")), colour = "red",lwd=3) +
  geom_text(data = arrow, aes(x = NMDS1, y = NMDS2, label = ef),
            size = 5)

#add site points
qq<-gg+geom_point(data=keep,aes(x=NMDS1,y=NMDS2,col=factor(mic_prod_abund_cat)),size=2)+ #note size outside aes so no legend
  geom_segment(data =arrow,
               aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")), colour = "red",lwd=3) +
  geom_text(data = arrow, aes(x = NMDS1, y = NMDS2, label = ef),
            size = 5) 
qq

###output plot
jpeg(here("output/nms_grp_mic_prod_abund_cat.jpeg")) 
qq
dev.off()


qq<-gg1 +
  geom_segment(data =arrow,
               aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")), colour = "red",lwd=3) +
  geom_text(data = arrow, aes(x = NMDS1, y = NMDS2, label = ef),
            size = 5, nudge_x=c(.4,.2,0), nudge_y = c(-.07,0,.05))
qq

```

* 3d plot of scores

```{r 3d, include=FALSE, eval=FALSE}
ordirgl(nms07,col=grp2)
```

* biplot of envfit  #https://stackoverflow.com/questions/14711470/plotting-envfit-vectors-vegan-package-in-ggplot2

```{r plot_scores, include=FALSE}
#sites
sites <- as.data.frame(scores(nms07, display = "sites"))
#add toxin detect data & WSA9
sites<-cbind(sites,select(env,WSA_ECO9,cyl_det,mic_det,sax_det,tox_detects))

#get envfit scores
ef.scrs <- as.data.frame(scores(ef07, display = "vectors"))
ef.scrs <- cbind(ef.scrs, ef = rownames(ef.scrs))


scale<-6
show1<-filter(ef.scrs,ef=='mic_det' | ef=='sax_det' | ef=='cyl_det')%>%
      mutate(NMDS1=NMDS1*scale,NMDS2=NMDS2*scale)

show2<-filter(ef.scrs,ef=='microcystin' | ef=='saxitoxin' | ef=='cylindrospermopsin')%>%
      mutate(NMDS1=NMDS1*scale,NMDS2=NMDS2*scale)



ggplot(sites) +
  geom_point(mapping = aes(x = NMDS1, y = NMDS2, colour = grp2)) +
  coord_fixed() + ## need aspect ratio of 1!
  geom_segment(data =arrow,
               aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")), colour = "orange",lwd=2) +
  geom_text(data = arrow1, aes(x = NMDS1, y = NMDS2, label = ef),
            size = 5)


ggplot(spp) +
  geom_point(mapping = aes(x = NMDS1, y = NMDS2, colour=microcystin,size=2)) +
  coord_fixed() + ## need aspect ratio of 1!
  geom_segment(data =arrow,
               aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")), colour = "orange",lwd=2) +
  geom_text(data = arrow1, aes(x = NMDS1, y = NMDS2, label = ef),
            size = 5)

plot(nms07,disp='sites',type='n')
ordihull(nms07,grp2,col=c('blue','orange'),lwd=2)
#ordiellipse(nms07,grp2,col=1:3,kind='ehull')
#ordiellipse(nms07,grp2,col=1:3,draw="polygon")
#ordispider(nms07,grp2,col=1:3,label=TRUE)
points(cyl$NMDS1,cyl$NMDS2,col='cyan',pch=16,cex=2)
points(mic$NMDS1,mic$NMDS2,col='blue',pch=16,cex=2)
points(sax$NMDS1,sax$NMDS2,col='green',pch=16,cex=2)
#points(nms07, disp="sites", pch=21, col="red", bg="yellow", cex=1.3)


```

* draw some box plots based on lake group

```{r boxplot, include=FALSE}
boxplot(log(env$mic_prod_abund)~grp2)
boxplot(log(env$cyl_prod_abund)~grp2)
boxplot(log(env$sax_prod_abund)~grp2)
boxplot(log(env$cyano_abund)~grp2)
boxplot(log(env$microcystin)~grp2)

summary(env$microcystin[grp2==1])
summary(env$microcystin[grp2==2])
table(grp2,env$mic_prod_abund_cat)



```

* random forest analysis of env variables to predict grp2

```{r rf_data, include=FALSE}
env_rf<-cbind(group2=factor(grp2,label=c('one','two')),env)
env_rf<-env_rf[complete.cases(env_rf),]
predictors<-as.data.frame(select(env_rf,-group2,-site_id,-DATE_COL))
predictors[sapply(predictors, is.character)]<-lapply(predictors[sapply(predictors, is.character)],as.factor)#covert _cat to factor
lake_group<-factor(env_rf$group2)
```

```{r rf_anal, include=FALSE, eval=FALSE}
#random forest analysis
set.seed(1956)
rf_grp2_env07<-randomForest(predictors,lake_group,ntree=10000,importance=TRUE,proximity=TRUE)
save(rf_grp2_env07,file=here("data/rf_grp2_env07.rda"))
```

```{r rf_presnt, include=FALSE}
#variable importance
#https://www.r-bloggers.com/variable-importance-plot-and-variable-selection/
load(here("data/rf_grp2_env07.rda"))
head(rf_grp2_env07$importance)
varImpPlot(rf_grp2_env07,type=2) #mean decrease in node impurity
varImpPlot(rf_grp2_env07,type=1) #mean decrease in accuracy

partialPlot(rf_grp2_env07,predictors,'WSA_ECO9')
partialPlot(rf_grp2_env07,predictors,'cyano_abund')
```

* random forest analysis of phyto variables to predict grp2

```{r rfphyto_anal, include=FALSE, eval=FALSE}
lake_group1<-as.factor(grp2) 
set.seed(1956)
rf_grp2_phyto07<-randomForest(phyto07[,-1],lake_group1,ntree=10000,importance=TRUE,proximity=TRUE)
save(rf_grp2_phyto07,file=here("data/rf_grp2_phyto07.rda"))
```

```{r rfphyto_present, include=FALSE}
load(here("data/rf_grp2_phyto07.rda"))
varImpPlot(rf_grp2_phyto07,type=2) #mean decrease in node impurity
varImpPlot(rf_grp2_phyto07,type=1) #mean decrease in accuracy

```

* Indicator Species Analysis

```{r ind_spp, include=FALSE}
#http://rfunctions.blogspot.com/2013/02/multivariate-analysis-indicator-value.html

ind_spp<-labdsv::indval(phyto07[,-1],grp2)

gr <- ind_spp$maxcls[ind_spp$pval<=0.05]
iv <- ind_spp$indcls[ind_spp$pval<=0.05]
pv <- ind_spp$pval[ind_spp$pval<=0.05]
fr <- apply(phyto07[,-1]>0, 2, sum)[ind_spp$pval<=0.05]
indvalsummary <- data.frame(group=gr, indval=iv, pvalue=pv, freq=fr)
indvalsummary <- indvalsummary[order(indvalsummary$group, -indvalsummary$indval),]

# Let us see the results
indvalsummary

#what about the toxins
inner_join(mutate(indvalsummary,genus=rownames(indvalsummary),tox))

```

library(vegan)
library(ggplot2)
library(grid)

data(dune)

# calculate distance for NMDS
NMDS.log<-log1p(dune)
set.seed(42)
sol <- metaMDS(NMDS.log)

scrs <- as.data.frame(scores(scores, display = "sites"))
scrs <- cbind(scrs, Group = c("hi","hi","hi","md","lo","hi","hi","lo","md","md",
                              "lo","lo","hi","lo","hi","md","md","lo","hi","lo"))

set.seed(123)
vf <- envfit(sol, NMDS.log, perm = 999)

vf

#So the r2 data is used to scale the values in columns NMDS1 and NMDS2. The final plot is produced with:

spp.scrs <- as.data.frame(scores(vf, display = "vectors"))
spp.scrs <- cbind(spp.scrs, Species = rownames(spp.scrs))

p <- ggplot(scrs) +
  geom_point(mapping = aes(x = NMDS1, y = NMDS2, colour = Group)) +
  coord_fixed() + ## need aspect ratio of 1!
  geom_segment(data = spp.scrs,
               aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")), colour = "grey") +
  geom_text(data = spp.scrs, aes(x = NMDS1, y = NMDS2, label = Species),
            size = 3)


